Linting Linearity in Core/System FC

### 2023-01-16 report.tex
Typeset linearly typed lambda calculus

\newcommand{\llet}[2]{\mathsf{let}\ #1\ \mathsf{in}\ #2}

@@

To understand how linear types are defined and used in practice, we present two
examples of anonymous functions, try to understand how they could be disregarded
as incorrect, and work our way up to linear types from them. The first receives a
file handle as an argument, ignores it, and returns $\star$ (read unit), which
is equivalent to C's
\textbf{void}.
% TODO: Better code formatting
\[
    \lambda h.~\textrm{return}~\star;
\]

And, similarly, the second receives a file handle and closes it
twice. Note that \textbf{close} has type $\textbf{Handle} \to \star$, that is, it's a
function that receives a handle and returns unit.
\[
    \lambda h.~\textrm{close}~h;~\textrm{close}~h;
\]

Both the above functions share the same type ($\textbf{Handle} \to \star$),
under a simple type system such as C's they are type correct, and both have
erroneous behaviours. The first forgets to close the handle and the second
closes it twice. Our goal is now to reach a type system that rejects these
two programs.

The key observation is that the argument of the function must be linear, and we
can augment the common function type to account for the linearity of its
argument. We take the function type $A \to B$ and replace the arrow with the
$\lolli$
%
\footnote{Since linear types are born from a
correspondence\cite{curry-howard} (the Curry-Howard isomrphism) with linear
logic\cite{linear-logic}, we borrow the $\lolli$ symbol and other linear logic
connectives to describe linear types.}
%
operator to denote a function that uses its argument exactly once: $A \lolli B$.
% Linear types are a powerful idea because they allow us to statically reason
% about resources in our program. A so called linear resource

% For example, to define a ...

% Linear types are a powerful idea

In order to give well defined semantics to a linear type system, we present a
linearly typed lambda calculus which is a very simple language with linear
types...

\[
  \begin{array}{lcl}
    s,t & ::= & \star\\
        & \mid & t \lolli s\\
        & \mid & t \oplus s\\
        & \mid & t \tensor s\\
        & \mid & t \with s\\
        & \mid & \bang t\\
   \end{array}
\]

>>>>

\newcommand{\llet}[2]{\mathsf{let}~#1~\mathsf{in}~#2}
\newcommand{\ccase}[2]{\mathsf{case}~#1~\mathsf{of}~#2}

@@

To understand how linear types are defined and used in practice, we present two
examples of anonymous functions that receive as an argument a handle to work
with and are responsible for closing it, we try to understand how the examples
could be disregarded as incorrect, and work our way up to linear types from
them. The first function ignores the received file handle and returns $\star$
(read unit), which is equivalent to C's \textbf{void}.
% TODO: Better code formatting
\begin{mathpar}
    \lambda h.~\textrm{return}~\star;
    \and
    \lambda h.~\textrm{close}~h;~\textrm{close}~h;
\end{mathpar}
% TODO: Equation references?

Ignoring the file handle which should have been closed by the function makes the
first function incorrect. Similarly, the second function receives the file handle and
closes it twice, which is incorrect not because it falls short of the
specification, but rather because the program will crash while executing it.
%
Additionally, both functions share the same type, $\textbf{Handle} \to \star$, i.e.
a function that takes a \textbf{Handle} and returns $\star$. The second function also
shares this type because \textbf{close} has type $\textbf{Handle} \to \star$.
%
Under a simple type system such as C's both functions are type correct (the
compiler will happily succeed), but both have erroneous behaviours. The first
forgets to close the handle and the second closes it twice. Our goal is now to
reach a type system that rejects these two programs.

The key observation to invalidating these programs is to focus on the function
type going between \textbf{Handle} and $\star$ and augment it to indicate that
\emph{the argument must be used exactly once}, or, in other words, that the
argument of the function must be linear. We take the function type $A \to B$
and replace the function arrow ($\to$) with the linear function arrow ($\lolli$)
%
\footnote{Since linear types are born from a correspondence with linear
logic\cite{linear-logic} (the Curry-Howard isomorphism\cite{curry-howard}), we
borrow the $\lolli$ symbol and other linear logic connectives to describe linear
types.}
%
operator to denote a function that uses its argument exactly once: $A \lolli B$.
%
Providing the more restrictive linear function signature $\textbf{Handle} \lolli
\star$ to the example programs would make both of them fail to typecheck because
they do not satisfy the linearity specification that the function argument
should only be used exactly once.

% Linear types are a powerful idea because they allow us to statically reason
% about resources in our program. A so called linear resource ?

In order to further give well defined semantics to a linear type system, we
present a linearly typed lambda calculus, a very simple language with linear
types, by defining what are syntatically valid programs through the grammar and
what programs are well typed through the typing rules.

An expression in the linearly typed lambda calculus can be $\star$...
A type in the linearly typed lambda calculus can be $\star$...

\begin{mathpar}
  \begin{array}{lcl}
    s,t & ::= & \star\\
        & \mid & t \lolli s\\
        & \mid & t \oplus s\\
        & \mid & t \tensor s\\
        & \mid & t \with s\\
        & \mid & \bang t\\
   \end{array}

\and

  \begin{array}{lcl}
      M,N & ::= & \star \mid \llet{\star = M}{N} \\
        & \mid & u \\
        & \mid & \lambda u. M \mid M~N\\
        & \mid & \textrm{inl}~M \mid \textrm{inr}~M \mid
        \ccase{M}{\textrm{inl}~u_1 \to N_1 \mid \textrm{inr}~u_2 \to N_2}\\
        & \mid & M \tensor N \mid \llet{u_1 \tensor u_2 = M}{N}\\
        & \mid & M \with N \mid \textrm{fst}~M \mid \textrm{snd}~M \\
        & \mid & \bang M \mid \llet{!u = M}{N} \\
   \end{array}
\end{mathpar}

\begin{mathpar}
    \infer*[right=($\tensor I$)]
    {\Delta \vdash M : A \and \Delta' \vdash N : B}
    {\Delta , \Delta' \vdash (M \tensor N) : A \tensor B}

    \and
\end{mathpar}

\begin{mathpar}
    \infer*[right=($\tensor E$)]
    {\Delta \vdash M : A \tensor B \and \Delta', u{:}A, v{:}B \vdash N : C}
    {\Delta , \Delta' \vdash \llet{u \tensor v}{N} : C}
\end{mathpar}

\begin{mathpar}
    \infer*[right=($\lolli I$)]
    {\Delta , u{:}A \vdash M : B}
    {\Delta \vdash \lambda u. M : A \lolli B}

    \and

    \infer*[right=($\lolli E$)]
    {\Delta \vdash M : A \lolli B \and \Delta' \vdash N : A}
    {\Delta, \Delta' \vdash M~N : B}
\end{mathpar}

\begin{mathpar}
    \infer*[right=($\with I$)]
    {\Delta \vdash M : A \and \Delta \vdash N : B}
    {\Delta \vdash M \with N : A \with B}

    \and

    \infer*[right=($\with E_L$)]
    {\Delta \vdash M : A \with B}
    {\Delta \vdash \textrm{fst}~M : A}

    \and

    \infer*[right=($\with E_R$)]
    {\Delta \vdash M : A \with B}
    {\Delta \vdash \textrm{snd}~M : B}
\end{mathpar}

\begin{mathpar}
    \infer*[right=($\oplus I_L$)]
    {\Delta \vdash M : A}
    {\Delta \vdash \textrm{inl}~M : A \oplus B}

    \and

    \infer*[right=($\oplus I_R$)]
    {\Delta \vdash M : B}
    {\Delta \vdash \textrm{inr}~M : A \oplus B}

    \and

    \infer*[right=($\oplus E$)]
    {\Delta \vdash M : A \oplus B \and \Delta', w_1{:}A \vdash N_1 : C \and
    \Delta', w_2{:}B \vdash N_2 : C}
    {\Delta, \Delta' \vdash \ccase{M}{\textrm{inl}~w_1 \to N_1 \mid \textrm{inr}~w_2 \to N_2} : C}
\end{mathpar}

\begin{mathpar}
    \infer*[right=($\star I$)]
    { }
    {\Delta \vdash \star : \star}

    \and

    \infer*[right=($\star E$)]
    {\Delta \vdash M : \star \and \Delta' \vdash N : B}
    {\Delta, \Delta' \vdash \llet{\star = M}{N} : B}

\end{mathpar}

\begin{mathpar}
    \infer*[right=($\bang I$)]
    {\Gamma ; \cdot \vdash M : A}
    {\Gamma ; \cdot \vdash !M : !A}

    \and

    \infer*[right=($\bang E$)]
    {\Gamma ; \Delta \vdash M : \bang A \and \Gamma, u{:}A ; \Delta' \vdash N : C}
    {\Gamma ; \Delta, \Delta' \vdash \llet{!u = M}{N} : C}
\end{mathpar}
